---
title: "Redlining in Current Environmental Injustices"
author: "Kate Becker"
date: "2023-12-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE)
```


# Overview 
The term "environmental injustice" is a word that encompasses our past, present, and future and it represents years of systemic oppression, marginalization, and inequality. Redlining, fore example, was banned 50 years ago but it continues to hurt minorities today. Throughout redlining's history, local zoning officials worked with businesses to place polluting operations such as industrial plants, major roadways and shipping ports around marginalized communities. In terms of housing, it categorically denied access to mortgages not just to individuals but to whole neighborhoods. Socioeconomic factors and redlining are actively being studied today but this analysis will study redlining but from a biodiversity lens. 



For more information regarding the workflow, data, and reproducibility of this repository please see the Github link below
Github: https://github.com/kateebeckerr/Redlining_Injustices


# Data Descriptors

1. EJ Screen: Data from the United States Environmental Protection Agency's Environmental Justice Screening and Mapping Tool. This data is publicly accessible in order to be more transparent about how they consider environmental justice(EJ) in their work, to assist stakeholders in making informed decisions about pursuing EJ, and to create a common starting point between the agency and the public when looking at issues in EJ. This data provides environmental and demographic information for the US and census tract and block groups but in this particular analysis, only the block group data will be explored. 

2. Mapping Inequality: This data comes from the Mapping Inequality: Redlining in New Deal America project. This is an online mapping tool that enables you to explore pre-existing redlining maps for almost every major American City as well as is history of raical and ethnic discrimination in housing policy. As we will see in this analysis,neighborhoods were given HOLC grades. Those that they deemed "best" and safe investments were given a grade of A and colored green. Those that were deemed "hazardous" were given a grade of "D" and colored red. The map of Los Angeles was selected for this analysis. 

3. Biodiversity: The Global Biodiversity Information Facility is a international network and data infrastructure funded by the worlds government that aims to provides free and open access to biodiversity data. The data used in this analysis is bird observations from 2021 and onward. 


## Relevant Libraries and Setting Path
```{r}
library(tidyverse)
library(sf)
library(tmap)
library(spData)
library(spDataLarge)
library(terra)
library(dplyr)
library(ggplot2)
library(raster)
library(maps)
library(spData)
library(grid)
library(viridis)
library(tmap)
library(tmaptools)

rm(list = ls())
```

## Data Read In

### LA EJ Screen 
```{r}
ejscreen <- st_read("./data/EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb/") 
#filter for LA county
LA <- ejscreen %>%
  filter(CNTY_NAME %in% c("Los Angeles County")) #filter for Los Angeles County
```
### Biodiversity Data
```{r}
biodiversity <- st_read("/Users/katebecker/Documents/Bren/Fall_Q/EDS_223/final/Redlining/Redlining_Injustices/data/gbif-birds-LA") 
```
### LA Redlining
```{r}
LA_redlining <- st_read("https://dsl.richmond.edu/panorama/redlining/static/citiesData/CALosAngeles1939/geojson.json") %>%
  st_make_valid() #makes a geometry valid by fixing any topological errors or inconsistencies in the geometry 
```

## Data Exploration 

Visualizing Variables of Interest 
```{r}
plot(LA["ID"])
plot(LA_redlining["grade"])
```

## Analysis 

#### Data Cleaning
In order to study the effects of wastewater discharge on census block groups, the census block groups that are above the 95th percentile of wastewater discharge (P_PWDIS) should be selected for and a centroid sould be added.

```{r}
LA_ejscreen_centroids <- st_centroid(LA) %>% #computes the centroid of each geometry in the LA spatial object
  filter(P_PWDIS > 95) #filters the centroids for census block groups above 95th percentile 

census_95 <- LA %>% 
  filter(P_PWDIS >= 95) #filters LA spatial object to include only geometries where P_PWDIS is greater than or equal to 95

census_95 <- st_centroid(census_95) #calculates the centroids of the geometries in census_95

coordLA <- st_bbox(LA) #extracted the coordinates of LA to use them in the following map 

print(coordLA) 
```

## Map
The effects of wastewater discharge on the census block groups 
```{r}
LA_waste_1 <- LA %>% #using LA spatial object
  ggplot() +
  geom_sf(data = LA, aes(fill = P_PWDIS))+ #add polygons representing wastewater discharge areas 
  geom_sf(data = census_95, color = "red", shape = 5, size = 0.5) + #adds red centroid dots to plot 
  scale_fill_gradient(low = "lightyellow", high = "darkorange", name = "Percentile for Wastewater discharge") +
  coord_sf(
    ylim = c(3980000, 4139895) #add x and y limits
  ) +
  ggtitle("Wastewater in Los Angeles County") #adds a title

LA_waste_1
```


### Low income communities 

Communities with less than 5% of the population considered low income (LOWINCPCT)

```{r}
low_income_5 <- LA %>%
  filter(LOWINCPCT < 0.05)  #filters data to include rows where the value in LOWINCPCT is less than 0.05
  
percent_low <- (nrow(low_income_5) / nrow(LA)) * 100 #calculates %

print(percent_low)
```


### Particulate Matter and Superfund Approximity 

Those above the 80th percentile for Particulate Matter 2.5 AND  
above the 80th percentile for Superfund proximity will be analyzed 

```{r}
has_data <- LA %>% #remvoes all NAs
  filter(!is.na(P_PM25)) %>%
  filter(!is.na(P_PNPL))

LA %>% 
  filter(P_PM25 > 80) %>% #filters for particualte matter and superfund approximity over 80%
  filter(P_PNPL > 80) %>%
  nrow()/nrow(has_data)*100 #calculates percentage
```


###Spatial Features 

```{r}
coordLA_red <- st_bbox(LA_redlining) #finds the bounding box coordinates to use for mapping in the next step 
print(coordLA_red)

crs1 <- st_crs(LA)
crs2 <- st_crs(LA_redlining)

LAred_transformed <- st_transform(LA_redlining, crs1) #transforms CRS' so they match 

LA_LAredline <- st_intersects(LAred_transformed, LA)

```

## Map: Historical Redlining Colored by HOLC Grade
```{r}
LA_LAred <- LA %>%
  ggplot() +
  geom_sf(data = LA) +
  geom_sf(data = LA_redlining, aes(fill = grade)) +
  ggtitle("Redlining in Los Angeles by HOLC Grade") +
  coord_sf(
    ylim = c(3980000, 4139895)
  )

print(LA_LAred)

```

### Census block groups that fall within areas with HOLC grades
```{r}
#change CRS so it matches
LA <- st_transform(LA, crs= st_crs(LA_redlining))

binary_list <- st_intersects(x = LA, y = LA_redlining) # returns binary predicate list
logical_intersect = lengths(binary_list) > 0 # create logical of which items do intersect
redlining_cbg <- LA[logical_intersect, ] # filter census block groups based on logicals

# the number of census block groups
print(paste("There are", nrow(redlining_cbg), "census block groups falling within areas with HOLC grades"))
```

Summarize current conditions by finding the mean of the following variables:
- % low income.  
- percentile for particulate Matter 2.5.  
- percentile for low life expectancy.  
- percentile for air toxics cancer risk
```{r}
LA_Grade <- st_transform(LA, crs = st_crs(LA_redlining)) #transforms geometry

LA_red <- st_intersection(LA_redlining, LA_Grade) #performs an intersection

#Analyzing the intersection of two spatial datasets and summarizing variables within each grade variable 
LA_red %>%
  group_by(grade) %>%
  summarise(lowincpct = mean(LOWINCPCT, na.rm = TRUE),
            pm25 = mean(P_PM25, na.rm = TRUE),
            lifeexppct = mean(P_LIFEEXPPCT, na.rm = TRUE),
            cancer = mean(P_CANCER, na.rm = TRUE)) 
```

 
### For bird observations from 2022 that fall within neighborhoods with HOLC grads the percent of observations within each redlining categories is calculated
```{r}
bio_22 <- biodiversity  %>%
  filter(year == 2022) # filter to 2022

# transform data to match CRS
biodiversity_transform <- st_transform(bio_22, crs = st_crs(LA_redlining))
```


## Map: The Legacy of Redlining in Biodiversity Observations 
```{r}
#joined data
biodiv_data <- st_join(x = biodiversity_transform, y = LA_redlining, join = st_within, left = FALSE)

#Calculates summary of the counts and percentages of features in each HOLC grade fro the join data 
biodiv_summary <- biodiv_data %>% 
  st_set_geometry(NULL) %>%
  group_by(grade) %>%                  
  summarise(count = n()) %>%      
  mutate(percentage = (count / sum(count))*100 )


ggplot(data = biodiv_summary) + #Creates a bar chart visualizing % observations by HOLC grade
  geom_bar(aes(x = grade, y = percentage, fill = grade), stat = "identity") +
  labs(x = "HOLC grade", y = "Percentage of observations") +
  scale_fill_discrete() +
  ggtitle("Redlining in Biodiversity Observations")
```






